### 2024-09-11 Article/main.tex
Graphs as functor categories

\section{Set-Theoretic Definition of Graphs}

@@

\subsection{Simple Graphs}

@@

\begin{definition}[Graph Homomorphism]
    Given two graphs $G=(V_G, E_G, s_G, t_G)$ and $H=(V_H, E_H, s_H, t_H)$, a \emph{graph homomorphism} $h: G \rightarrow H$ is a pair of functions $h = (h_V: V_G \rightarrow V_H, h_E: E_G \rightarrow E_H)$ such that
    \[
        h_V \circ s_G = s_H \circ h_E
    \]
    and
    \[
        h_V \circ t_G = t_H \circ h_E
    \]
    that is, a structure preserving map.
    Given two graph homomorphisms $h = (h_V, h_E): G \rightarrow H$, $k = (k_V, k_E): H \rightarrow I$, the homomorphism $ k \circ h = (k_V \circ h_V, k_E \circ h_E): G \rightarrow I$ is the \emph{composite} of $h$ and $k$.
\end{definition}

@@

\begin{example}
    The initial object in $\mathbf{Graph}$ is the empty graph, i.e., the graph with an empty set of vertices and an empty set of edges. The initial object instead is the graph with exactly one node and a single edge from that node to itself.
\end{example}

@@

These examples show how some kind of limits are computed among graphs, but we have already said something similar about categories of presheaves (Observation~\ref{obs:limits_in_presh}). In fact, it is possible to show the equivalence between the category of graphs and $[\bullet \rightrightarrows \bullet , \Set]$. The easy way of computing depends on the fact that in a category of presheaves, limits are computed objectwise [citation needed], as we implicitly have done in the previous examples.

@@

\begin{remark}
    It is possible to define an equivalence relation $\sim \ \subseteq S\times S$ as a surjective function mapping each $s \in S$ to a partition of $S$ in which each element is equivalent according to $\sim$, that is, its \emph{equivalence class}. Formally, an equivalence relation $\sim \ \subseteq S\times S$ is fully described by a function $f_\sim : S \rightarrow \{[s]_\sim \mid s \in S \}$, where $[s]_\sim = \{t \in S \mid t \sim s\}$, defined by $f_{\sim}(s) = [s]_\sim$. 
\end{remark}

@@

In the category of graphs with equivalences universal constructions are easy to obtain. 
The terminal object in $\mathbf{EqGrph}$ is the graph with only one node, only one edge and with the single node equivalent only to itself, while the initial object is the graph with no nodes and no edges.
Given two graphs with equivalences $\mathcal{G} = (G, \sim_G) $ and $\mathcal{H} = (H, \sim_{H})$, one can construct the product graph $\mathcal{G \times H} = (G \times H, \sim_{G \times H})$, together with the two projections $\pi_G$ and $\pi_H$,  where $\sim_{G \times H}$ is the least equivalence relation on $G \times H$ such that, if $w_1$ and $w_2$ are nodes of $\mathcal{G \times H}$, we have $w_1 \sim_{G \times H} w_2$ if $\pi_G(w_1) \sim_G \pi_G(w_2)$ and $\pi_H(w_1) \sim_H \pi_H(w_2)$.
An equalizer in $\mathbf{EqGrph}$ for a pair of morphisms $f, g: \mathcal{G \rightarrow H}$ is exactly the same as in $\mathbf{Graph}$, namely, $(\mathcal{E} = (E, \sim_E), e)$, where $E$ is the greatest subgraph of the underlying graph of $\mathcal{G}$ on which $f$ and $g$ are equal, together with the inclusion $e: \mathcal{E \rightarrow G}$. The equivalence relation is preserved ($\sim_E$ is $\sim_G$ restricted to the set of vertices of $E$).
Coequalizers are computed in the same way. Computing 
The two previous examples shows that the category $\mathbf{EqGrph}$ has both equalizer and finite products, so, by Theorem~\ref{th:limit}, we can conclude that $\mathbf{EqGrph}$ has finite limits.

>>>>

\section{Set-Theoretic Definitions of Graphs}

@@

\subsection{Graphs}

@@

\begin{definition}\label{def:graph_hom}[Graph Homomorphism]
    Given two graphs $G=(V_G, E_G, s_G, t_G)$ and $H=(V_H, E_H, s_H, t_H)$, a \emph{graph homomorphism} $h: G \rightarrow H$ is a pair of functions $h = (h_V: V_G \rightarrow V_H, h_E: E_G \rightarrow E_H)$ such that
    \[
        h_V \circ s_G = s_H \circ h_E
    \]
    and
    \[
        h_V \circ t_G = t_H \circ h_E
    \]
    that is, a structure preserving map.
    Given two graph homomorphisms $h = (h_V, h_E): G \rightarrow H$, $k = (k_V, k_E): H \rightarrow I$, the homomorphism $ k \circ h = (k_V \circ h_V, k_E \circ h_E): G \rightarrow I$ is the \emph{composite} of $h$ and $k$.
\end{definition}

@@

\begin{example}\label{ex:in_term_in_graph}
    The initial object in $\mathbf{Graph}$ is the empty graph, i.e., the graph with an empty set of vertices and an empty set of edges. The initial object instead is the graph with exactly one node and a single edge from that node to itself.
\end{example}

@@

@@

\begin{remark}\label{rem:eq_as_surj}
    It is possible to define an equivalence relation $\sim \ \subseteq S\times S$ as a surjective function mapping each $s \in S$ to a partition of $S$ in which each element is equivalent according to $\sim$, that is, its \emph{equivalence class}. Formally, an equivalence relation $\sim \ \subseteq S\times S$ is fully described by a function $f_\sim : S \rightarrow \{[s]_\sim \mid s \in S \}$, where $[s]_\sim = \{t \in S \mid t \sim s\}$, defined by $f_{\sim}(s) = [s]_\sim$. 
\end{remark}

@@

%TODO: RIGUARDARE QUESTA PARTE!!!

In the category of graphs with equivalences, universal constructions are easy to obtain. 
The terminal object in $\mathbf{EqGrph}$ is the graph with only one node, only one edge and with the single node equivalent only to itself, while the initial object is the graph with no nodes and no edges.
Given two graphs with equivalences $\mathcal{G} = (G, \sim_G) $ and $\mathcal{H} = (H, \sim_{H})$, one can construct the product graph $\mathcal{G \times H} = (G \times H, \sim_{G \times H})$, together with the two projections $\pi_G$ and $\pi_H$,  where $\sim_{G \times H}$ is the least equivalence relation on $G \times H$ such that, if $w_1$ and $w_2$ are nodes of $\mathcal{G \times H}$, we have $w_1 \sim_{G \times H} w_2$ if $\pi_G(w_1) \sim_G \pi_G(w_2)$ and $\pi_H(w_1) \sim_H \pi_H(w_2)$.
An equalizer in $\mathbf{EqGrph}$ for a pair of morphisms $f, g: \mathcal{G \rightarrow H}$ is exactly the same as in $\mathbf{Graph}$, namely, $(\mathcal{E} = (E, \sim_E), e)$, where $E$ is the greatest subgraph of the underlying graph of $\mathcal{G}$ on which $f$ and $g$ are equal, together with the inclusion $e: \mathcal{E \rightarrow G}$. The equivalence relation is preserved ($\sim_E$ is $\sim_G$ restricted to the set of vertices of $E$).

Previous examples show that the category $\mathbf{EqGrph}$ has both (c)equalizer and finite (co)products, so, by Theorem~\ref{th:limit}, we can conclude that $\mathbf{EqGrph}$ has finite limits and colimits.

%TODO: EGGS

\section{Graphs as Functors}

Set-theoretical definitions of graphs are straightforward, but a further level of abstraction allows us to highlight some interesting properties of these objects. We can indeed consider a graph as a functor from a category of ``primitive objects'' (e.g., the object representing edges and the object representing vertices) onto a category which provides a representation of them. This concept is perfectly represented by categories of presheaves, where an object is nothing than a way to interpret objects of a category as sets. An example will make clear what we are talking about.

Considering the category of graphs (as defined in Definition~\ref{def:cat_of_graph}), it is possible to show that 
$$\mathbf{Graph} \cong [E \mathrel{\mathop{\rightrightarrows}^{s}_{t}} V , \Set]$$
where an object $G$ is a functor, having $G(V)$ as the set of vertices, $G(E)$ as the set of edges, $G(s)$ and $G(t)$ as the source and target functions. A morphisms $\eta: G \dot\rightarrow H$ is a natural trasformation, and the commutativity of the diagram below is given by definition of graph morphisms (Definition~\ref{def:graph_hom})
$$
    \begin{tikzcd}
        G(E) \arrow[r, "{\eta_E}"] \arrow[d, "G(s)"swap] & H(E) \arrow[d, "H(s)"] \\
        G(V) \arrow[r, "{\eta_V}"swap] & H(V)
    \end{tikzcd}
$$
(and the same for $t$).

Example~\ref{ex:in_term_in_graph} and following, in which limits and colimits are computed component-wise, are a concrete instance of Observation~\ref{obs:limits_in_presh}.

More complex objects are graphs with equivalence. We can think of $\mathbf{EqGrph}$ as a subcategory of the category $[V \rightrightarrows E \rightarrow Q, \Set]$, in which each object $\mathcal{G}$ is a functor mapping the arrow $q: E \rightarrow Q$ onto an epimorphism (i.e., a surjective function). The component $\mathcal{G}(q)$ here is intended to be the function discussed in Remark~\ref{rem:eq_as_surj}.
